---
title: "HW 26 - Practice one way ANOVA"
format: html
editor: visual
---

# Practice 1-way ANOVA

### Set up and get the data

```{r}
rm(list=ls())
library(here)
library(tidyverse)
library(ggfortify)
```

```{r}
peas <- read.csv(here("Data", "s_r_pea_sections.csv"))
```

Look at the data

```{r}
glimpse(peas)
```

Change data type

```{r}
peas$Treatment <- as.factor(peas$Treatment)
levels(peas$Treatment)
```

### Question:

Is there a difference in the mean pea section lengths between the four treatments? How do each of the other three treatments compare to the control treatment?

### Plot

Make a box and whiskers plot to look at the data

```{r}
ggplot(peas, aes(x=Treatment, y=pea_section_length_ou))+
  geom_boxplot()+
  theme_bw()+
  coord_flip()
```

#### First interpretations

Control has a much higher mean pea section length than the other treatments, with 2% sucrose following.

Estimations of means:

-   Control = 71

-   2 percent sucrose = 64

    -   diff from control \~ 7

-   2 percent glucose = 59

    -   diff from control \~ 12

-   2 percent fructose = 56

    -   diff from control \~ 15

-   1 percent fructose 1 percent glucose = 56

    -   diff from control \~ 15

Degrees of freedom = (5-1) = 4

## Run the ANOVA

### Construct a model

```{r}
model.peas <- lm(pea_section_length_ou ~ Treatment, data = peas)
```

### Test Assumptions

```{r}
autoplot(model.peas)
```

#### Interpretation:

Seems good - residuals vs fitted line is pretty flat, normal Q-Q is close to the sloped line, no major points with leverage

### Run the ANOVA

```{r}
anova(model.peas)
```

#### Results:

Df = 4

F-val = 49.368

P-val is small

-   There is a significant difference in the mean pea section lengths between the treatments (Table 1. F\[4, 45\] = 49.368, p\<\<0.001).

### Treatment Contrasts

```{r}
summary(model.peas)
```

#### Interpretation:

The reference treatment that everything's being compared to here is the '1 percent fructose 1 percent glucose', which is not what we want to compare...

This is saying that all the treatments have higher means than the '1 percent fructose 1 percent glucose' treatment. The '2 percent sucrose' and 'control' treatments have a significant difference in mean.

## Now the Tukey HSD Test

activate the multcomp package

```{r}
library(multcomp)
```

```{r}
Tukey.test <- glht(model.peas, linfct = mcp(Treatment = "Tukey"))
```

```{r}
summary(Tukey.test)
```

```{r}
cld(Tukey.test)
```

### Interpretation

This test shows that the control treatment has a significant difference in the mean pea section length from the other treatments. '2 percent sucrose' is also significantly different from the other treatments. The other three treatments ('1 percent fructose 1 percent glucose,' '2 percent fructose,' and '2 percent glucose') do not have a significant difference in their means.
