---
title: "Homework 27"
format: html
editor: visual
---

Read through the code in this assignment. You will begin contributing to the code on line 36. Your tasks will be to:

-   Create a dataframe with just the complete cases

-   Review the 1-way ANOVA we ran during class with the full data set

-   Conduct a 2nd 1-way ANOVA using just the data set with complete cases and compare the results to the first analysis.

-   Conduct a linear regression to assess the relationship b/w Growth_rate_mm and Dried_mass_g, first with the complete data set and second with the complete cases data set and compare the results of the 2 analyses.

## Set up our workspace and read in the data

```{r}
rm(list = ls())
library(tidyverse)
library(ggfortify)
library(here)
```

Read in the data.

```{r}
df <- read.csv(here("Data/learn-about-NA.csv"))
str(df)
df$Treatment <- as.factor(df$Treatment)
```

## Dealing with NAs in data analysis

Deciding how to treat NA values in your data can have critical impacts on your data analysis.

Let's look at an example.

Start by creating a data frame from df called df_All_NA, which includes the entire dataset including all NA values.

```{r}
df_All_NA <- df
```

Now use one of the methods we discussed in class on 13 November to create a second dataframe called `df_No_NA` that is just the complete cases of df.

```{r}
df_No_NA <- df[complete.cases(df), ] 
```

What if we wanted to test the hypothesis that there is no effect of fertilizer treatment on plant growth rate?

In class on Wednesday 13 November we ran a 1-way ANOVA on the complete data set. Let's review it here:

### First ANOVA using all the data

#### Plot and Guess

```{r}
ggplot(df_All_NA, aes(Treatment, Growth_rate_mm))+
  geom_boxplot()+
  geom_point()+
  theme_bw()
```

We decided that there was likely an effect of treatment on growth rate, but we weren't sure if both levels of the fertilizer treatment would differ from the control.

#### Create a model:

```{r}
grow.mod<-lm(Growth_rate_mm ~ Treatment, data = df_All_NA)
```

#### Check our assumptions:

```{r}
autoplot(grow.mod)
```

Looks pretty good.

#### Interpret results:

```{r}
anova(grow.mod)
```

Practice your interpretation.

Why are there 2 DF for the Treatment effect?

Answer: Because there are 3 levels of our Treatment variable: Control, Level 1, and Level 2. Number of groups - 1 = degrees of freedom therefore 3 - 1 = 2.

Why are there 46 DF for the Residuals? Remember that we determined that there were 49 values in the `Growth_rate_mm` variable using `sum(!is.na(variable))` like this:

```{r}
sum(!is.na(df$Growth_rate_mm))
```

So **there are only 49 data points used in the analysis**, even though the dataframe is 60 rows long!!!

49 - 3 = 46 = degrees of freedom for the residuals.

```{r}
summary(grow.mod)
```

Practice your interpretation:

Does adding 1 dose of fertilizer change the growth rate significantly from the Control?

-   *No, the p-value is 0.0812 which is not significant.*

Does adding 2 doses of fertilizer change the growth rate significantly from the Control?

-   *Yes, the p-value is 0.0108 which is significant is using the cut off of 0.05 for the p-value.*

We won't worry about re-plotting our data right now.

### Second Attempt: Complete Cases

Your turn. Run the 1-way ANOVA again, this time with only the dataset that has complete cases and that we called `df_No_NA`

#### Plot and Guess

```{r}
ggplot(df_No_NA, aes(Treatment, Growth_rate_mm))+
  geom_boxplot()+
  geom_point()+
  theme_bw()
```

What do you think?

*This seems similar to our first graph, so I would guess that either one or both of the levels will differ from the control, but they won't differ from each other significantly.*

#### Create a model:

```{r}
grow.mod.no.NA <-lm(Growth_rate_mm ~ Treatment, data = df_No_NA)
```

#### Check our assumptions:

```{r}
autoplot(grow.mod.no.NA)
```

What do you think? Same or different from our first 1-way ANOVA?

*This one is a little bit different in that the scale-location line dips more than the first ANOVA. The residuals vs leverage points are also a bit more spread out.*

#### Interpret results:

```{r}
anova(grow.mod.no.NA)
```

Practice your interpretation.

Why are there 2 DF for the Treatment effect?

-   *There are 3 levels in the Treatment variable so 3-1=2 degrees of freedom*.

Why are there 33 DF for the Residuals?

-   *There are 36 observations being used in this analysis and 3 variables being used, so 36-3=33.*

How would you write a results statement?

*There is not a significant difference between the three treatments on growth rate (F\[2, 85.19\]=2.7906, p=0.0759).*

```{r}
summary(grow.mod.no.NA)
```

Practice your interpretation:

Does adding 1 dose of fertilizer change the growth rate significantly from the Control?

-   *No, the p-value is high.*

Does adding 2 doses of fertilizer change the growth rate signficantly from the Control?

-   *Depends on what is considered significant. Yes if using p\<0.05, but no if using p\<0.01.*

We won't worry about re-plotting our data right now.

## Linear Regression

Now conduct a linear regression to examine the relationship between `Growth_rate_mm` and `Dried_mass_g`, first with the complete data set and second with the complete cases data set and compare the results of the 2 analyses.

### First attempt: Full data set including NAs (use `df_All_NA`)

You know what to do here. Go for it.

#### Plot and guess

Use a scatterplot to visualize this data as both are continuous.

```{r}
ggplot(df_All_NA, aes(x=Dried_mass_g, y=Growth_rate_mm))+
  geom_point()+
  xlab("Dried mass (g)")+
  ylab("Growth rate (mm)")+
  theme_bw()

```

This graph has a positive slope, but the points are pretty widely distributed. I think there will be a relationship between the two variables but it won't have a very small p-value.

The growth rate ranges from about 5 to 22 (rise - 17)

The dried mass ranges from about 30 to 52 (run - 22)

So the slope is about 17/22= 0.7727273

#### Create model

```{r}
mod.reg.ALL <- lm(Dried_mass_g ~ Growth_rate_mm, data = df)
```

#### Check assumptions

```{r}
autoplot(mod.reg.ALL)
```

Don't worry if assumptions don't look perfect - focus on normality of the data.

-   *This is the case, data seems overall normally distributed but not perfect.*

#### Interpret model

Use `anova()` first before using `summary()`

```{r}
anova(mod.reg.ALL)
```

Why is there 1 degree of freedom for Growth_rate_mm and why are there 34 degrees of freedom for the Residuals?

*There are 34 degrees of freedom because it is the number of observations - the number of parameters. In this case, it says in the summary below that 24 observations were "deleted due to missingness" so from the 60 original observations we end up with 36. There are 2 parameters (intercept and dried_mass_g), so 36-2=34.*

```{r}
summary(mod.reg.ALL)
```

There is a significant relationship between dried mass and growth rate (F\[1,34\]=10.805, p=0.002356

### Second attempt: Limited data set with only complete cases (use `df_No_NA`)

#### Plot and guess

```{r}
ggplot(df_No_NA, aes(x=Dried_mass_g, y=Growth_rate_mm))+
  geom_point()+
  xlab("Dried mass (g)")+
  ylab("Growth rate (mm)")+
  theme_bw()

```

*This looks like the same graph.*

#### Create model

```{r}
mod.reg.NO <- lm(Dried_mass_g ~ Growth_rate_mm, data = df)
```

#### Check assumptions

```{r}
autoplot(mod.reg.NO)
```

Again, don't worry if the assumptions aren't perfect. How do they compare to the first analysis?

*They look just about the same.*

#### Interpret model

Use `anova()` first then `summary()`.

```{r}
anova(mod.reg.NO)
```

This is the same as the ANOVA table in the last test. There is a significant relationship between dried mass and growth rate. (F\[1,34\]=10.805, p=0.002356).

```{r}
summary(mod.reg.NO)
```

Answer the following questions:

1.  How many degrees of freedom are there for Growth_rate_mm? Why?
    1.  *There is 1 degree of freedom. Because the degrees of freedom for the intercept are always set at 1 (I think)*
2.  How many degrees of freedom are there for Residuals? Why?
    1.  *36-2=34 which is the number of observations (36) minus the number of parameters (2)*
3.  How are the degrees of freedom in this second model different than the first? What does the answer tell you about the role of NA values in the regression analysis?
    1.  *These are the same degrees of freedom as in the first model. This might say that R automatically takes out NA values when doing a model?*
