---
title: "Homework 16 due Oct 9"
format: html
editor: visual
---

# Reinforcing GSWR Chp 3 and in class work

Set up

```{r}
rm(list=ls())
library(tidyverse) #includes dplyer
library(here)
```

Import data and create dataframe

```{r}
cover <- read.csv(here("Data", "donnerville_percent_cover.csv"))
```

Examine data set

```{r}
glimpse(cover)
str(cover)
```

## Change column names

Changing column name of cover_type to be uppercase and match other names better

```{r}
colnames(cover)[8] <- "Cover.type" 
#colnames(data frame)[the column in question] <- "the new name"
```

## Change data types

Make Site, Plot type, and Cover type into factor variables

```{r}
cover$Site <- as.factor(cover$Site)
cover$Plot.type <- as.factor(cover$Plot.type)
cover$Cover.type <- as.factor(cover$Cover.type)

glimpse(cover)
```

Make Date into a date variable

```{r}
cover$Date <- as.Date(cover$Date)
```

Never mind the above, need to use **lubridate** instead

```{r}
cover$Date <- dmy(cover$Date) #dmy means day, month, year - how the date is entered in the table
# R then coverted the dates to year, month, day format
```

## Error checking

Need to edit Site (currently has 7 levels, should have 3), Cover type (currently has 6 levels, should have 3)

```{r}
#For Donnerville 1

levels(cover$Site)
fix1 <- which(cover$Site %in% c("DONNERVILLE - 1", "donnerville 1", "Donnerville 1")) #identifies the rows that need repair (all those with these names) and assigns them to the fix1 object

cover$Site[fix1] <- "Donnerville 1" #apply the object to change the values

cover$Site <-factor(cover$Site) #tells R to relist the levels of factor b/c have changed

levels(cover$Site)
```

```{r}
#Now for Donnerville 2

fix2 <- which(cover$Site %in% c("donnerville 2", "Donnerville 2"))

cover$Site[fix2] <- "Donnerville 2"

cover$Site <- factor(cover$Site)

levels(cover$Site)
```

```{r}
#Now for Donnerville 3
#Try a diff way from class using logical operator ==

fix3 <- which(cover$Site == "donnerville-3")

cover$Site[fix3] <- "Donnerville 3"

cover$Site <- factor(cover$Site)

levels(cover$Site) 
#Success!
```

Now to edit Cover type column

```{r}
levels(cover$Cover.type) 
```

```{r}
#Start with Bare

fix4 <- which(cover$Cover.type == "bare")

cover$Cover.type[fix4] <- "Bare"

cover$Cover.type <- factor(cover$Cover.type)
levels(cover$Cover.type)
```

```{r}
#Now Bryophytes

fix5 <- which(cover$Cover.type == "bryophtes")

cover$Cover.type[fix5] <- "Bryophytes"

cover$Cover.type <- factor(cover$Cover.type)
levels(cover$Cover.type)
```

```{r}
#Lastly, Vascular

fix6 <- which(cover$Cover.type == "vasc ")

cover$Cover.type[fix6] <- "Vascular"

cover$Cover.type <- factor(cover$Cover.type)
levels(cover$Cover.type)
```

## 

## Determine average percent cover by cover type (bare, bryophytes, vascular plants) for Donnerville 1

```{r}
summarise(
  group_by(cover, Site, Cover.type), meanPercent = mean(Percent)
)

# What's going on in each line
# summarise according to the following directions
# group_by (the data frame, the columns to sort output by), mean Percent is telling what math to do and what column to do it with
```

In Donnerville 1, the average cover type for Bare is 77.5%, for Bryophytes is 2.5%, for Vascular is 20%

Testing to see if I can change the output to only show Donnerville 1

```{r}
summarise(
  group_by(cover, "Donnerville 1", Cover.type), meanPercent = mean(Percent)
)
#If I replace Site column with just "Donnerville 1" will it give same output but just the Donnerville 1 rows?
```

Seemed to work but gave me different numbers... not sure why

## Determine the average percent vascular plant cover across all sites

```{r}
cover %>% 
  group_by(Cover.type)%>% 
  summarise(meanPercent = mean(Percent)) 

# line by line what's going on:
# use the cover data frame
# group it by the variable cover type (group into Bare, Bryophytes, Vascular)
# take the average/mean of each of these cover types (across all sites b/c don't have anything about site in this)

```

The average percent of vascular plant cover across all sites is 16.667%.

Testing to see if I can get the output to just be vascular plants (not the other two cover types as well)

```{r}
cover %>% 
  group_by(Cover.type)%>%
  group_by("Vascular")%>%
  summarise(meanPercent = mean(Percent)) 
```

Gave me just vascular but it definitely didn't work to give the correct average percent cover.
