---
title: "Biostats - Homework 20"
format: html
editor: visual
---

### Introduction

The objective of this assignment is to have you continue practicing the skills you have learned in Chapters 3 and 4 of Getting Started with R and reinforce the ideas we discussed in class regarding the Weissgerber et al. paper. Please complete all of the following steps:

1.  Write the code to set up your workspace.

```{r}
rm(list=ls())
library(tidyverse)
library(here)
```

2.  Read in the dataset called soil_data.csv and name it "soil."

```{r}
soil <- read.csv(here("Data", "soil_data.csv"))
```

3.  Create a histogram for the SOM_percent column in the dataframe. Accept the default number of bins.

```{r}
ggplot(soil, aes(x = SOM_percent))+
  geom_histogram()
```

How would you describe the mean_SOM data based on viewing the histogram? Write your answer below:

*The mean is skewed to the right (meaning the tail is longer on the right side).*

4.  Create a new dataframe called SOM by calculating for each forest the mean soil organic matter and the standard deviation of soil organic matter. Make sure to name variables you create "mean_SOM" and "sd_SOM" so that the next code chunk I give you works.

```{r}
SOM <- soil |>
  group_by(Forest) |>
  summarise(mean_SOM = mean(SOM_percent), sd_SOM = sd(SOM_percent))
# checked back to HW 15 - remember this is the format to use for summarizing and means!

```

5.  Now run the following code chunk which will create a bar plot of mean soil organic matter per forest, with standard deviation as error bars.

```{r}
ggplot(SOM, aes(Forest, mean_SOM))+
  geom_errorbar(aes(ymax = mean_SOM + sd_SOM, ymin = mean_SOM - sd_SOM), width = 0.2)+
  geom_col(fill = "white", color = "black")+
  ylab("Mean percent soil organic matter")+
  xlab("Forest name")
```

Based on what you see in the bar chart, how would you describe the pattern of SOM between the 4 different forests? Write your answer here:

Now create a boxplot of the SOM_percent data. Use Forest as the x-axis variable and overlay the point data on top of the boxplot. Make the points blue and somewhat transparent. Call the x-axis "Forest name" and the y-axis "Percent soil organic matter" (hint: See homework 18...).

```{r}


```

Based on what you see in the box plot, write out your answers the following questions:

-   how would you describe the pattern of SOM between the 4 different forests?

-   What do you notice that is different from how the data were displayed in the bar plot?

-   Which plot type is more informative? Explain your answer.
