---
title: "Homework 21"
format: html
editor: visual
---

# GSWR Chapter 5 first half

Set up

```{r}
rm(list=ls())
library(tidyverse)
library(here)
```

# Chi-square contingency table analysis

## Notes and set up

-   Test of association among two or more categorical variables

-   Have set of observations and can classify them by more than one categorical variable with levels

-   tests against null hypothesis of no association

-   not directional results

Ladybird data: frequency of black and red ladybirds in industrial and rural habitats (two grouping variables each with two levels) - are dark morphs more likely in dark industrial habitats

Question: "is some feature of the habitat associated with the frequencies of different color morphs?"

```{r}
#Read in data
lady <- read.csv(here("Data", "ladybirds.csv"))
```

```{r}
#Look at it
glimpse(lady)
#Need to change from character data to factor
```

```{r}
#Need to change from character data to factor
lady$Habitat <- as.factor(lady$Habitat)
lady$Site <- as.factor(lady$Site)
lady$colour <- as.factor(lady$colour)
#Re-ran glimpse to verify after each one
```

## Organizing the data for plotting and analysis

-   Need to calculate totals - four numbers - sums of observations of each color in each habitat

```{r}
totals <- lady |>
  group_by(Habitat, colour) |>
  summarise(total.number = sum(number))
```

Make a **bar chart** to summarize (only time recommended for summarizing data)

```{r}
ggplot(totals, aes(x = Habitat, y = total.number, fill = )) +
  geom_bar(stat = 'identity', position = 'dodge')+
  scale_fill_manual(values = c(black = "black", red = "red"))

#Using a bar chart here because it's not data points being used (that would have a distribution) it's actual values. There also is a 0 (it's ratio data).

# Use fill = __ when geometry can be filled
# Stat = 'identity' tells to not calculate anything
# position = 'dodge' tells to put the two bars in each group next to each other so it doesn't end up stacked
# Use scale_fill_manual to customize fill colors
```

The bar graph suggests we will reject the null hyp (that they are equally distributed) because they don't appear equally distributed in the habitats by color.

## Making the x2 (chi squared) test

Need to make a matrix of the total counts (diff from a data frame) using the xtabs() function

```{r}
lady.mat <- xtabs(number ~ Habitat + colour, data = lady)

#"Cross-tabulate the number column of counts in the lady data frame by the Habitat and colour variables" --> similar to group_by and summarise earlier but end result is a matrix not a data frame
```

Now chi square test

```{r}
chisq.test(lady.mat)
```

Results are a very low p-value so can confidently reject the null hypothesis aka it isn't a chance result and there is some association between the variables

*"We tested the hypothesis that there is an association between color morphs of ladybirds and industrial and rural habitats. Ladybird color morphs are not equally distributed in the two habitats (x2 = 19.1, df = 1, p \< 0.001), with black morphs being more frequent in industrial habitats and red morphs more frequent in rural habitats"*

```{r}
#See the mechanics of the chi squ test:
lady.chi <- chisq.test(lady.mat)
names(lady.chi)

lady.chi$expected
```

# Two sample t-test

## Notes and set up

-   A comparison of the means of two groups of numeric values

-   Use when sample sizes in each group are small

-   Makes assumptions: normal distribution, variances are equal

**Ozone example:**

Test if there's a difference in mean ozone concentrations between the east and west gardens.

```{r}
rm(list=ls())
library(tidyverse)
library(here)
```

```{r}
ozone <- read.csv(here("Data", "ozone.csv"))
```

```{r}
glimpse(ozone)

ozone$Garden.location <- as.factor(ozone$Garden.location)
ozone$Garden.ID <- as.factor(ozone$Garden.ID)
```

Now use histogram and facets to plot the data

```{r}
ggplot(ozone, aes(x = Ozone))+
  geom_histogram(binwidth = 10)+
  facet_wrap(~ Garden.location, ncol = 1)+
  theme_bw()
```

The peaks of the histograms don't overlap so may reject null hyp, but there is overlap in the graph so need stats to be sure.

Calculate means and st. errors of ozone levels in each location:

```{r}
Ozone.mean <- ozone|>
  group_by(Garden.location) |>
  summarise(mean.ozone = mean(Ozone), stdev.ozone = sd(Ozone))
```

## The two sample T-test analysis

```{r}
#Do a t-test
t.test(Ozone ~ Garden.location, data = ozone)

# "Do ozone levels vary as a function of garden location?" is indicated by ~
```

Output line by line:

1.  statistical test performed
2.  Data used
3.  t-test stat, degrees freedom, p-val
4.  gives the alternative hypothesis - true difference in means does not equal 0 (this would mean they are the same and could not reject the null) - so we can reject the null
5.  95% confidence interval = interval around the difference between the two means. (would be 0 if the means were the same), doesn't include 0 so we can reject the null and conclude probably difference between east and west
6.  Means in each group

Note - Welch = method allows one of the assumptions fo t-test - equal variance - to be relaxed

If need to make a formal test with test for variance could use (get F-stat)

```{r}
var.test(Ozone ~ Garden.location, data=ozone)
```
