---
title: "Homework 28 - chp 8"
format: html
editor: visual
---

# GSWR Chp 8

This chapter focuses on making nice plots with ggplot.

## Preparation

Normal set up

```{r}
rm(list=ls())
library(tidyverse)
library(here)
library(ggplot2) #as said to do in the book
```

Need to get gridExtra package - installed using the console

```{r}
library(gridExtra) #load the package
```

Read in the data

```{r}
compensation <- read.csv(here("Data", "compensation.csv"))
```

### First graphs

Start by making a scatterplot with Fruit and Root. Then make a box and whisker plot with grazing and fruit and assign both of these to an object. This is like what we did in chp 4.

```{r}
#BASE scatterplot
eg_scatter <- 
  ggplot(data=compensation, aes(x=Root, y=Fruit))+
  geom_point()

#BASE box-and-whiskers plot
eg_box <- 
  ggplot(data=compensation, aes(x=Grazing, y=Fruit))+
  geom_boxplot()
```

Because they're saved, we can use and add to the figure.

```{r}
eg_scatter + theme_bw() #Adds the remove background function to the scatterplot saved
```

Now use the gridExtra package for some things. The gridExtra package is a helper package for placing more than one ggplot2 figure on the same page.

```{r}
grid.arrange(eg_scatter, eg_box, nrow=1) 
#puts both the scatterplot and boxplot together on the same page!
```

## 8.3 - What you may want to customize

Use `scale_()` functions for parts that are tied to the x and y variables (defined by `aes()`)

Use `theme()` for parts of graph not directly determined by `aes()` like gridlines and text formatting.

## 8.4 - Axis labels, axis limits, and annotation

Recap - use `xlab` and `ylab` to change axis titles –\> only tell R what words to use, what to show

```{r}
eg_scatter + xlab("Root Biomass") + ylab("Fruit Production") #Change the axis names
```

`ggtitle` if need a title - rare and not recommended

```{r}
eg_scatter+ggtitle("My SUPERB title")
```

Can combine with `labs` the last two code chunks into:

```{r}
eg_scatter + labs(title="My useless title", x="Root Biomass", y="Fruit Production")
```

Changing the x and y axes... `xlim()` and `ylim()`

```{r}
eg_scatter + xlim(0, 20) + ylim(0, 140)
```

### Putting custom text inside the figure

Using the `annotate()` function

Need to specify: what kind of annotation (text), where to put annotation with the coordinate system of graph, what to write/the label

```{r}
eg_scatter +
  annotate("text", x = c(6,8), y = c(105, 25), label=c("Text up here...", "...and text down here"))
#Adds text (or other) to the figure - very neat
```

## 8.5 - Scales

`scale_()` functions are tied to the variables being plotted (x and y)

```{r}
eg_scatter + scale_x_continuous(limits = c(4,11), breaks = 4:11)
#limits adjusts range of x-axis
#breaks specifies integer values where want tick marks
```

### **Changing colors and fills**

Change colors based on levels of a grouping variable.

```{r}
ggplot(data=compensation, aes(x=Root, y=Fruit, colour=Grazing))+
  geom_point()+
  scale_colour_manual(values = c(Grazed = "brown", Ungrazed = "green"))
```

### **Transform the axis of the plot**

log-transform the y-axis of a plot to manage non-linearity or emphasize extent of variation

The trans = \_\_ can allocate many transformations to either or both of the x and y axes

```{r}
eg_box +
  scale_y_continuous(breaks = seq(from = 10, to = 150, by = 20, trans = "log10"))

# used seq() to generate vector of breakpoints for the ticks, then asked for log10 transformation
```

## 8.6 - The theme

Use the `theme()` function to adjust the non-aesthetic pieces of a graph

### panels and gridlines

First graph - get rid of grey background and minor gridlines, generate light blue major gridlines

```{r}
eg_scatter +
  theme(
    panel.background = element_rect(fill = NA, colour = "black"), 
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "lightblue")
  )

```

Breakdown of above code chunk:

-   panel has theme elements that can be adjusted and correspond to figure features (here its background, grid minor, grid major)

-   `element()` specifies arguments for the panel group geometric components (rectangles and lines)

    -   Customize these with fill = \_\_ and colour = \_\_

-   `element_blank()` shuts down all rendering of a `panel.()` component - here its to get rid of the minor grid lines

### axis titles and tick marks

Adjust the color and size of x-axis title and angle of the x-axis tick labels - helpful for text-based labels

```{r}
eg_box +
  theme(
    axis.title.x = element_text(colour = "cornflowerblue",
    size = rel(2)),
    axis.text.x = element_text(angle = 45, size = 13, vjust = 0.5)
  )
```

Breakdown:

-   `axis.()` class of attributes centers on manipulating text elements `element_text()` using arguments of size = \_\_ or size = rel(2) which increases size relative to default

-   vjust = \_\_ takes values b/w 0 and 1 to vertically adjust the labels

### discrete-axis customizations

Discrete axes delineate different groups of data according to the values of a categorical variables - represented by a factor or character vector in R - small number of unique values.

Use `scale_x_discrete()` or `scale_y_discrete()` to customize labeling of groups

Alter the names of the levels on graphs:

```{r}
eg_box + scale_x_discrete(limits = c("Ungrazed", "Grazed"),
                          labels = c("Control", "Grazed"))
```

### customizing legends/keys

Generate a "clean" key:

```{r}
ggplot(compensation, aes(x=Root, y=Fruit, colour=Grazing))+
  geom_point()+
  theme(legend.key = element_rect(fill=NA)) #this line removes the borders of the key
```

Have no key:

```{r}
ggplot(compensation, aes(x=Root, y=Fruit, colour=Grazing))+
  geom_point()+
  theme(legend.position = "none") #this line removes the key
```

All done with Chp 8!
