---
title: "Homework-26-ANOVA"
format: html
editor: visual
execute:
  output: TRUE
---

### Introduction

Let's practice running a 1-way ANOVA and interpreting the results. The data are on pea plant root lengths, measured in optical units (a microscope length unit), for plants grown on different growth media.

#### Set things up

You know what to do here

```{r}
rm(list = ls())
library(ggfortify)
library(multcomp)
library(tidyverse)
library(here)
```

Now pull in the dataset

```{r}
peas <- read.csv(here("Data/s_r_pea_sections.csv"))
```

Treatment should be our factor variable.

```{r}
peas$Treatment <- as.factor(peas$Treatment)
levels(peas$Treatment)
```

The factor levels look good.

Part 1: Is there a difference in mean pea lengths?

Part 2: If there is a difference, for which growth media does the difference exist?

### Plot the data

Let's start with a histogram of the data

```{r}
ggplot(peas, aes(pea_section_length_ou))+
  geom_histogram()+
  facet_wrap(~Treatment, ncol = 1)
```

Now a box plot

```{r}
ggplot(peas, aes(Treatment, pea_section_length_ou))+
  geom_boxplot()+
  geom_jitter(width = 0.1)+
  coord_flip()
```

Guesstimate the results:

Peas grown in the control medium will be significantly longer than other pea plants.

Control: mean c. 72 ou 2% sucrose: mean c. 64 2% glucose: mean c. 58 2% fructose: mean c. 56 1% fructose 1% glucose: mean c. 56

### Build the model

Let's make the model

```{r}
pea_mod<- lm(pea_section_length_ou ~ Treatment, data = peas)
```

### Check model assumptions

```{r}
autoplot(pea_mod, smooth.colour = NA)
```

Basically looks ok

### Interpret results

```{r}
anova(pea_mod)
```

There is a signficant effect of growth treatment on pea plant length.

And now summary

```{r}
summary(pea_mod)
```

Because 1% fructose 1% glucose is first alphabetically, the contrasts show us the difference in mean between the that and each other growth treatment.

Based on that, we see that: Control plants grew signficantly longer than the 1% glucose 1% fructose treatment, as did plants in the 2% sucrose treatment. Neither 2% fructose or 2% glucose were different from 1% fructose 1% glucose.

It would be nice if we had a method for multiple comparison of means. We do, but we need to use different pacakges.

We are going to do what are called "post hoc" tests of difference in means. To do so, we need the `multcomp` package. We turned it on with `library(multcomp)` at the top of our file.

Let's conduct a Tukey's HSD (honestly significant differences) test.

```{r}
phc1<- glht(pea_mod, linfct = mcp(Treatment = "Tukey"))
summary(phc1)
```

According to this result: There is no difference in mean pea length for peas grown in 2% fructose vs. 1% f 1% s, nor is there a differentce in mean pea length for plants grown in 2% glucose vs. 1% f 1% s. There is also no difference in mean pea length for peas grown in 2% glucose vs. those grown in 2% fructose. All other groups were signficiantly different from one another.

We can also ask for a "compact letter display" of these results

```{r}
cld(phc1)
```

Or we can use the `emmeans()` function from the `emmeans` package

```{r}
library(emmeans)
ph2 <- emmeans(pea_mod, specs = "Treatment")
summary(ph2)
```

Then use the `pairs()` function to run the Tukey test

```{r}
pairs(ph2)
```
